---
title: "hw02"
author: "wade"
date: "September 23, 2017"
output: html_document
---

Welcome to my Stat 545 hw02 page. In this homework assignment, I will be exploring the gapminder dataset by using the dplyr package that we learned about during class.

First, I will load the packages I need:
```{r}
require(tidyverse)
require(gapminder)

```

The gapminder dataset is <b>rectangular</b> data, which means, as far as I can surmise, that if you look at all of the data at once it literally is shaped like a rectangle.

Next, let's "smell test" the data to get a feel for what it is like.

```{r}

dim(gapminder)
nrow(gapminder)
ncol(gapminder)
is.data.frame(gapminder) #It appears that this is indeed a dataframe
is.vector(gapminder) # but not a vector
is.matrix(gapminder) # nor a matrix
is.list(gapminder) # It is also a list, because a dataframe is a list of lists

class(gapminder) # This command is probably more simple to use, but I like when R yells TRUE or                     FALSE at me. According to this function, the dataset consists these classes:                     "tbl_df," "tbl" and "data.frame."

str(gapminder) # There are 1704 observation of 6 variables
                #these variables have data types: 2 are factors, 2 are integers, and 2 are                      numeric
summary(gapminder)

```
As shown, there are several ways to get information about the parameters of a dataset.
At this time, I do not see the value in all of these functions (there is a lot of redundance),
but I'm sure this will change as I become more familiar with R.

Now I will explore <u> one categorical variable and one numeric variable.</u>

```{r}
levels(gapminder$continent)
# Interestingly, there are 5 levels of this variable:
# "Africa"   "Americas" "Asia"     "Europe"   "Oceania"

summary(gapminder$continent)
  # Africa Americas     Asia   Europe  Oceania 
  #    624      300      396      360       24

```

Let's see if we can show this in graphical form:
```{r continents, echo=FALSE}
pie(summary(gapminder$continent), main="Continents")
```

I should note that pie charts are not the best to use (they aren't as informative as barplots). But they're fun.

Now we can move on to a numeric variable.

```{r}
summary(gapminder$lifeExp)
  #  Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  # 23.60   48.20   60.71   59.47   70.85   82.60 

```

```{r lifeExp1, echo=FALSE}
hist((gapminder$lifeExp), main="Histogram", xlab="Life Expectancy")
```

```{r lifeExp2, echo=FALSE}
boxplot((gapminder$lifeExp), main="Boxplot", xlab="Life Expectancy")
```

From these graphs we can see the basic spread of the variable "lifeExp."

Now, let's explore this variable in more depth using ggplot2.
```{r lifeExp3, echo=FALSE}
ggplot(gapminder, aes(x=gdpPercap, y=lifeExp)) +
  geom_point(col="salmon", alpha=.25)
```

Let's try something new and fancy.
```{r lifeExp4, echo=FALSE}
g<- ggplot(gapminder, aes(x=continent, y=lifeExp))
g+ geom_jitter(alpha=.5, aes(color=continent), position=position_jitter(width=.2))
```

This variable follows a Poisson distribution:
```{r gdpPercap, echo=FALSE}
ggplot(gapminder, aes(gdpPercap)) + geom_density(color="slategrey")
```


Next, I will play with select, filter, and piping.

What was the life expectancy for the continent <b>Oceania</b> in 1962?

```{r}
filter(gapminder, continent=="Oceania", year==1962)

```

```{r}

filter(gapminder, lifeExp<70) %>% select(country,pop, lifeExp) %>%tail(10)

```

Piping is more simple than I expected, and is a huge quality of life improvement.

```{r}
filter(gapminder, continent=="Africa", pop>100000000) %>% select(country, pop, year) %>% head(3)
```

Extra stuff from the hw page:
```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```

Something is wrong with the above code! It would be more appropriate to use the %in% command, which tells R to look for elements within the concatonated set.

```{r}
filter(gapminder, country %in% c("Rwanda", "Afghanistan"))
```

This brings back twice as many values as the other code. It looks like R was doing something strange, like averaging the two countries together.

#Process

<li>This homework was simple (the in-class training was excellent, especially for dplyr), but it did take me a long time to complete.</li>
<li> I used my class notes a lot during this assignment </li>
<li> I also used some of the links from the Stat 545 website, including <a href="http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/">this one.</a> </li>




